Multiresolution Neural Field Estimation

### 2011-02-28 Multiresolution_framework/ltx/manuscript.tex
had a crack at tidying up the last bit of §4. Made two suggestions for cuts.

The state-space representation of the MRAIDE allows the use of the well known Expectation Maximization (EM) algorithm \cite{Dempster1977,Shumway2000} to infer both the kernel and the field from electrophysiological data. The EM algorithm, when used to in this context \cite{Dewar,Scerri,Freestone}, yields the maximum likelihood kernel estimate and the posterior distribution of the field over time. We use the standard EM algorithm for linear dynamic systems \cite{Gibsona2005,Roweis97} \mike{Roweis and Gharamani, 1997, A Unifying Review of Linear Guassian Models} and hence herein only describe aspects of the algorithm that are specific to MRA neural field equations. EM essentially iteratively finds increasingly tighter lower bounds on the likelihood of the kernel so that, at convergence, the maximum of the bound corresponds to the maximum of the likelihood function. For a multivariate linear dynamic system, the bound used is a quadratic of the form
\begin{equation}
 \mathcal Q\left(\boldsymbol \theta,\boldsymbol\theta'\right)=\beta+2T_s\left(\boldsymbol\upsilon_0-\boldsymbol\upsilon_1\right)\boldsymbol\theta-T_s\boldsymbol\theta^\top\boldsymbol\Upsilon\boldsymbol\theta,
\end{equation}
which has a maximum at
\begin{align}
 \boldsymbol \theta= \boldsymbol\Upsilon^{-\top}(\boldsymbol\upsilon_0^\top-\boldsymbol\upsilon_1^\top),
\end{align}
where
\begin{align}\label{eq:upsilon0}
 \boldsymbol\upsilon_0&=\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_0]_{i,j}[\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf U]_{j,i}\\ 
 \boldsymbol\upsilon_1&=\xi\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf U]_{j,i}\label{eq:upsilon1} \\
 \boldsymbol\Upsilon&=T_s\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\mathbf{U}^{\mathsf T} \boldsymbol\Lambda_{x}^{-1}\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf{U}]_{j,i}\label{eq:Upsilon}.
\end{align}
The $\left(i,j\right)^{th}$ element of the block matrix $\mathbf U_{n_x \times n_x}$ is given by
\begin{align}
\left[ \mathbf U\right] _{i,j}&=\int_{\boldsymbol \Omega}\left[\mu(r) \right]_i \left[\int_{\boldsymbol\Omega} \boldsymbol\mu\left(r'\right)\boldsymbol \lambda^\top \left(\mathbf{r-r'}\right) dr'\right]_{j:} dr.
\end{align}
where $[.]_{j:} $ denotes $j^{th}$ row of the matrix. The second terms in Eqs. \ref{eq:upsilon0}, \ref{eq:upsilon1}, and \ref{eq:Upsilon} can be computed as a once-off before the commencement of the EM iterations, which increases the speed of the M-step significantly compared to the method in \cite{Dewar2009}. $\Xi$-variables are calculated using the smoother outputs, i.e., state estimates, $\hat{\mathbf x}_t$ along with their covariance matrix, $\mathbf P_t$, and the cross-covariance matrix $\mathbf M_t$ using \cite{Gibsona2005}
\begin{align}\label{eq:Xivariables}
\boldsymbol\Xi_0&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \\
 \boldsymbol\Xi_1&=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right).
\end{align}
\dean{Number of iterations and stopping criteria?}The stopping criterion is usually associated  with  either the change in the parameter estimates or the log-likelihood variation \cite{McLachlan1997}.

>>>>

The state-space representation of the MRAIDE allows the use of the well known Expectation Maximization (EM) algorithm \cite{Dempster1977,Shumway2000} to infer both the kernel and the field from electrophysiological data. The EM algorithm, when used to in this context \cite{Dewar2009,Scerri2009,Freestone2011}, yields the maximum likelihood kernel estimate and the posterior distribution of the field over time. We use the standard EM algorithm for linear dynamic systems \cite{Gibsona2005,Roweis97} \mike{Roweis and Gharamani, 1997, A Unifying Review of Linear Guassian Models} and hence herein only describe aspects of the algorithm that are specific to MRA neural field equations. EM essentially finds increasingly tighter lower bounds on the likelihood of the kernel so that, at convergence, the maximum of the bound corresponds to the maximum of the likelihood function. For a multivariate linear dynamic system, the bound used is a quadratic of the form
\begin{equation}
 \mathcal Q\left(\boldsymbol \theta,\boldsymbol\theta'\right)=\beta+2T_s\left(\boldsymbol\upsilon_0-\boldsymbol\upsilon_1\right)\boldsymbol\theta-T_s\boldsymbol\theta^\top\boldsymbol\Upsilon\boldsymbol\theta,
\end{equation}
which has a maximum at
\begin{align}
 \boldsymbol \theta= \boldsymbol\Upsilon^{-\top}(\boldsymbol\upsilon_0^\top-\boldsymbol\upsilon_1^\top),
\end{align}
where
\begin{align}\label{eq:upsilon0}
 \boldsymbol\upsilon_0&=\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_0]_{i,j}[\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf U]_{j,i}\\ 
 \boldsymbol\upsilon_1&=\xi\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf U]_{j,i}\label{eq:upsilon1} \\
 \boldsymbol\Upsilon&=T_s\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\mathbf{U}^{\mathsf T} \boldsymbol\Lambda_{x}^{-1}\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf{U}]_{j,i}\label{eq:Upsilon}.
\end{align}
The $\left(i,j\right)^{th}$ element of the block matrix $\mathbf U_{n_x \times n_x}$ is given by
\begin{align}
\left[ \mathbf U\right] _{i,j}&=\int_{\boldsymbol \Omega}\left[\mu(r) \right]_i \left[\int_{\boldsymbol\Omega} \boldsymbol\mu\left(r'\right)\boldsymbol \lambda^\top \left(\mathbf{r-r'}\right) dr'\right]_{j:} dr.
\end{align}
where $[.]_{j:} $ denotes $j^{th}$ row of the matrix.\cut{The second terms in Eqs. \ref{eq:upsilon0}, \ref{eq:upsilon1}, and \ref{eq:Upsilon} can be computed as a once-off before the commencement of the EM iterations, which increases the speed of the M-step significantly compared to the method in \cite{Dewar2009}.} The matrices $\Xi_0$ and $\Xi_1$ are calculated using the Rauch Tung Streibel \cite{Rauch}(RTS) smoother outputs: state estimates, $\hat{\mathbf x}_t$, covariance, $\mathbf P_t=\mathrm{cov}(\mathbf{x_t})$, and cross-covariance matrix $\mathbf M_t=\mathrm{cov}(\mathbf{x}_t,\mathbf{x}_{t-1})$ 
\begin{align}\label{eq:Xivariables}
\boldsymbol\Xi_0&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \\
 \boldsymbol\Xi_1&=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right).
\end{align}

Hence, the algorithm has two distinct steps: the E-step which computes $\Xi_0$ and $\Xi_1$ based on the most recent parameter estimates using the RTS smoother, and the M-step which updates the paramter estimates by computing the (analytic) maximum of $Q(\theta,\theta')$. The EM algorithm iterates between these two steps until the parameter estimates converge; multiple restarts are required to avoid convergence on a local maxima.

\cut{\dean{Number of iterations and stopping criteria?}The stopping criterion is usually associated  with  either the change in the parameter estimates or the log-likelihood variation \cite{McLachlan1997}.}
